Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-04-08 papers/halfdisk/halfdisk.tex
readthrough

\title{Sparse spectral and $p$-finite element methods on disk-slices and trapeziums I: zero boundary conditions}
\author{Sheehan Olver, Ben Snowball}
%\date{}							% Activate to display a given date or no date

@@

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, we consider the solution of partial differential equations on the domain
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
where we require that either of the following scenarios hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and the weight function \(w_2\) on the interval $(\gamma, \delta)$ is an even function.
\end{enumerate}
We will explain where $w_2$ comes into play in Section \ref{Section:OPs}.

@@

Sparsity comes from expanding the domain and range of an operator  using different choices of $a$ and $b$. Where the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle  results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside quadrature rules to determine the entries.  Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}.

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

\begin{proposition}
Let \(w_1 : (\alpha,\beta) \: \to \R\), \(w_2 : (\gamma,\delta) \: \to \R\) be weight functions with \(\alpha,\beta,\gamma,\delta \in \R\), and let \(\rho \: : \: (\alpha,\beta) \: \to (0,\infty)\) be such that either:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w_2\) is an even function.
\end{enumerate}
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then the 2D polynomials defined on $\Omega$
$$
\Pnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k(\frac{y}{\rho(x)}) \qquad\hbox{for} \qquad 0 \le k \le n, \: n = 0,1,2,\dots
$$
are orthogonal polynomials with respect to the weight \(W(x,y) := w_1(x) \: w_2(\frac{y}{\rho(x)}) \) on $\Omega$. 
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

We can generalise this for our purposes in the following proposition. \bstodoinline{I have added in this proposition to explain how I think this generalises to the main domain shapes we are thinking about, including your triangle case.}
\begin{proposition}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $w^{(a,b,c)}_H(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align}
\begin{cases}
w^{(a,b,c)}_H(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{H_n^{(a,b,c)}\}$ are orthogonal with respect to $w^{(a,b,c)}_H(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

@@

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= w_H^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align}
\bstodoinline{Hope the following makes sense. The generalisation I have, and how I originally wrote/coded up the half-disk case use slightly different conventions regarding the parameters - would it be worth reframing the paper to match this generalisation, or is this fine?}
Note here the we have made the adjustment that $w^{(a,b,c)}_H(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite{magnus1995painleve}) and we introduce the notation \(\{H_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials \(\{\Pnkab\}\) given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}
 \Pnk^{(a,b)}(x,y) := H_{n-k}^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align}
W^{(a,b)}(x,y) = w_H^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align}

@@

\begin{lemma}
$\Pnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align}
x \Pnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: P_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: P_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: P_{n+1, k}^{(a,b)}(x, y), \\
y \Pnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: P_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: P_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: P_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: P_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: P_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: P_{n+1, k+1}^{(a,b)}(x, y),
\end{align}
for \((x,y) \in \Omega\), where
\begin{align}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)} \\
\alphaab_{n,k,2} &:= \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= \frac{\delta_{k-1}^{(b)}}{|| P_{n-1,k-1}^{(a,b)} ||_{W^{(a,b)}}^2} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{n-k}^{(a, b+k-\half)}>_{w_H^{(a, b+k+\half)}} \\
\betaab_{n,k,2} &:= \frac{\delta_{k}^{(b)}}{|| P_{n-1,k+1}^{(a,b)} ||_{W^{(a,b)}}^2} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{n-k-2}^{(a, b+k+\frac{3}{2})}>_{w_H^{(a, b+k+\frac{3}{2})}} \\
\betaab_{n,k,3} &:= \frac{\delta_{k-1}^{(b)}}{|| P_{n,k-1}^{(a,b)} ||_{W^{(a,b)}}^2} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{n-k+1}^{(a, b+k-\half)}>_{w_H^{(a, b+k+\half)}} \\
\betaab_{n,k,4} &:= \frac{\delta_{k}^{(b)}}{|| P_{n,k+1}^{(a,b)} ||_{W^{(a,b)}}^2} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{n-k-1}^{(a, b+k+\frac{3}{2})}>_{w_H^{(a, b+k+\frac{3}{2})}} \\
\betaab_{n,k,5} &:= \frac{\delta_{k-1}^{(b)}}{|| P_{n+1,k-1}^{(a,b)} ||_{W^{(a,b)}}^2} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{n-k+2}^{(a, b+k-\half)}>_{w_H^{(a, b+k+\half)}} \\
\betaab_{n,k,6} &:= \frac{\delta_{k}^{(b)}}{|| P_{n+1,k+1}^{(a,b)} ||_{W^{(a,b)}}^2} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{n-k}^{(a, b+k+\frac{3}{2})}>_{w_H^{(a, b+k+\frac{3}{2})}}. 
\end{align}
where $|| p(x,y) ||_{W^{(a,b)}}^2 := \int_\Omega p(x,y)^2 \: \Wab(x,y) \: d\mathbf{x}$ and $\ip<p(x), q(x)>_{w_H^{(a, b)}} := \int_\alpha^\beta p(x) \: q(x) \: x^a \: (1-x^2)^b \: dx$, i.e the 1D inner product using the $\{H^{(a,b)}_n\}$ family weight of $w_H^{(a, b)}$.

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align}
\bigPab_n := \begin{pmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{pmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\bigPab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bigPab(x,y) = x \: \bigPab(x,y), \quad J_y^{(a,b)} \: \bigPab(x,y) = y \: \bigPab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a functions expansion in the $\{\Pnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bigPab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bigPab(x,y)^\top {J_x^{(a,b)}}^\top \mathbf{f}$. In other words, ${J_x^{(a,b)}}^\top \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \to x \: f(x,y)$ in the  $\{\Pnkab\}$ basis.

$J_x^{(a,b)}, J_y^{(a,b)}$ have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber \\
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align}

@@

Combining each system in (\ref{eqn:jacobimatricesdefinition}) we can write the block-tridiagonal system
\begin{align}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bigPab(x,y)
=
\begin{pmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{pmatrix}.
\end{align}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv H_0^{(a,b)} \: P_0^{(b,b)}\), and for each $n = 0,1,2\dots$:
\begin{align}
A_n &:= \begin{pmatrix}
		A^x_n \\
		A^y_n \\
		A^z_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+3)} \quad (n \ne N), \quad
C_n := \begin{pmatrix}
		C^x_n \\
		C^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{pmatrix}
		B^x_n \\
		B^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{pmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align}

For each $n = 0,1,2\dots$ let $\Dnt$ be any matrix that is a left inverse of $A_n$, i.e. such that $\Dnt A_n = I_{n+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dnt$'s, we obtain a lower triangular system \cite[p78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots.
\end{cases}
\end{align}

@@

>>>>

\title{Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums}
\author{Sheehan Olver, Ben Snowball}
%\date{}							% Activate to display a given date or no date

@@

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, we consider the solution of partial differential equations on the domain
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
where we require that either of the following scenarios hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\).
\end{enumerate}

@@

Sparsity comes from expanding the domain and range of an operator  using different choices of $a$ and $b$. Whereas the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle  results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside quadrature rules to determine the entries.  Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk \sotodo{Add citation to this}, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

\begin{proposition}
Let \(w_1 : (\alpha,\beta) \: \to \R\), \(w_2 : (\gamma,\delta) \: \to \R\) be weight functions with \(\alpha,\beta,\gamma,\delta \in \R\), and let \(\rho \: : \: (\alpha,\beta) \: \to (0,\infty)\) be such that either:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w_2\) is an even function.
\end{enumerate}
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then the 2D polynomials defined on $\Omega$
$$
\Pnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)}) \qquad\hbox{for} \qquad 0 \le k \le n, \: n = 0,1,2,\dots
$$
are orthogonal polynomials with respect to the weight \(W(x,y) := w_1(x) \: w_2(\frac{y}{\rho(x)}) \) on $\Omega$. 
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

We can generalise this for our purposes in the following proposition.
\begin{proposition}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $w^{(a,b,c)}_H(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align}
\begin{cases}
w^{(a,b,c)}_H(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{H_n^{(a,b,c)}\}$ are orthogonal with respect to $w^{(a,b,c)}_H(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

@@

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= w_H^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align}
\bstodoinline{Hope the following makes sense. The generalisation I have, and how I originally wrote/coded up the half-disk case use slightly different conventions regarding the parameters - would it be worth reframing the paper to match this generalisation, or is this fine?}
\sotodoinline{Yes I think its fine} 
Note here the we have made the adjustment that $w^{(a,b,c)}_H(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite{magnus1995painleve}) and we introduce the notation \(\{H_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials\sotodo{We should probably use a different letter than $P$ to avoid confusion with Jacobi polynomials}
\ \(\{\Pnkab\}\) given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}\label{eq:diskpolys}
 \Pnk^{(a,b)}(x,y) := H_{n-k}^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align}
W^{(a,b)}(x,y) = w_H^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align}

@@

\begin{lemma}
$\Pnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align}
x \Pnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: P_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: P_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: P_{n+1, k}^{(a,b)}(x, y), \\
y \Pnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: P_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: P_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: P_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: P_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: P_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: P_{n+1, k+1}^{(a,b)}(x, y),
\end{align}
for \((x,y) \in \Omega\), where
\begin{align}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)} \\
\alphaab_{n,k,2} &:= \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= {\delta_{k-1}^{(b)}}{|| P_{n-1,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k}^{(a, b+k-1/2)}>_{w_H^{(a, b+k+1/2)}} \\
\betaab_{n,k,2} &:= {\delta_{k}^{(b)}}{|| P_{n-1,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k-2}^{(a, b+k+{3/2})}>_{w_H^{(a, b+k+{3/2})}} \\
\betaab_{n,k,3} &:= {\delta_{k-1}^{(b)}}{|| P_{n,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k+1}^{(a, b+k-1/2)}>_{w_H^{(a, b+k+1/2)}} \\
\betaab_{n,k,4} &:= {\delta_{k}^{(b)}}{|| P_{n,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k-1}^{(a, b+k+{3/2})}>_{w_H^{(a, b+k+{3/2})}} \\
\betaab_{n,k,5} &:= {\delta_{k-1}^{(b)}}{|| P_{n+1,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k+2}^{(a, b+k-1/2)}>_{w_H^{(a, b+k+1/2)}} \\
\betaab_{n,k,6} &:= {\delta_{k}^{(b)}}{|| P_{n+1,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<H_{n-k}^{(a, b+k+1/2)}, H_{n-k}^{(a, b+k+{3/2})}>_{w_H^{(a, b+k+{3/2})}}. 
\end{align}
where $|| p(x,y) ||_{W^{(a,b)}}^2 := \int_\Omega p(x,y)^2 \: \Wab(x,y) \: d\mathbf{x}$ and $\ip<p(x), q(x)>_{w_H^{(a, b)}} := \int_\alpha^\beta p(x) \: q(x) \: x^a \: (1-x^2)^b \: dx$, i.e the 1D inner product using the $\{H^{(a,b)}_n\}$ family weight of $w_H^{(a, b)}$.

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align}
\bigPab_n := \begin{pmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{pmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\bigPab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bigPab(x,y) = x \: \bigPab(x,y), \quad J_y^{(a,b)} \: \bigPab(x,y) = y \: \bigPab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a functions expansion in the $\{\Pnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bigPab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bigPab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \to x \: f(x,y)$ in the  $\{\Pnkab\}$ basis.

$J_x^{(a,b)}, J_y^{(a,b)}$ have the Banded-Block-Banded\sotodo{Define Banded-Block-sBanded} structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber \\
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align}

@@

Combining each system in (\ref{eqn:jacobimatricesdefinition}) we can write the block-tridiagonal system
\begin{align}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bigPab(x,y)
=
\begin{pmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{pmatrix}.
\end{align}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv H_0^{(a,b)} \: P_0^{(b,b)}\), and for each $n = 0,1,2\dots$:\sotodo{Remove $A^z_n$?}
\begin{align}
A_n &:= \begin{pmatrix}
		A^x_n \\
		A^y_n \\
		A^z_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+3)} \quad (n \ne N), \quad
C_n := \begin{pmatrix}
		C^x_n \\
		C^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{pmatrix}
		B^x_n \\
		B^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{pmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align}

For each $n = 0,1,2\dots$ let $\Dnt$ be any matrix that is a left inverse of $A_n$, i.e. such that $\Dnt A_n = I_{n+3}$\sotodo{n+1?}. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dnt$'s, we obtain a lower triangular system \cite[p78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots.
\end{cases}
\end{align}

@@

\sotodoinline{Rewrite in a more "paper" like tone}
